Slitasje og kjørespor

### 2023-01-30 DRAFT_slitasjeOgKjørespor.Rmd
feat: conditional paths; load ET map

```{r}
Sys.Date()
```

```{r setup, include=FALSE}
library(knitr)
library(sf)
library(tmap)
library(ggplot2)
library(ggpubr)
library(DT)
library(units)
library(eaTools)
library(tidyverse)
library(tidyr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

1.  Import [Nature type data set](#NTM) set (incl. GRUK) and [ANO data
    set](#ANO-import)

@@

6.  [Scale](#scale-slitasje-ind) the `slitasje` variable

@@

9.  Crop based on ecosystem delineation

10. Confirm relationship between infrastructure index and indicator
    values to justify the extrapolation

11. Calculate uncertainty?

12. Export indicator maps and extrapolated maps

## About the underlying data

### Representativity in time and space

### Original units

@@

### <!-- Text here -->

@@

```{r}
dir <- substr(getwd(), 1,2)
```

```{r, eval=F}
path <- ifelse(dir == "C:", 
               "P:/41201785_okologisk_tilstand_2022_2023/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb",
               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb")

naturetypes <- sf::st_read(dsn = path)
```

@@

```{r}
path <- ifelse(dir == "C:", 
               "P:/41201785_okologisk_tilstand_2022_2023/data/rds/naturtyper.rds",
               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/rds/naturtyper.rds")
naturetypes <- readRDS(path)
```

@@

```{r, fig.cap="7SE scores in the naturetype dataset"}
temp <- naturetypes_wide %>%
  group_by(var_7SE) %>%
  summarise(sum = n())

@@

```{r, fig.cap="7TK scores in the nature type dataset"}
temp <- naturetypes_wide %>%
  group_by(var_7TK) %>%
  summarise(sum = n())

@@

```{r slitasje-naturetype, fig.cap="Slitasje scores."}
temp <- naturetypes_wide %>%
  group_by(slitasje) %>%
  summarise(sum = n())

@@

```{r, fig.cap="Counting the number of cases where 7TK or 7SE ws defining of the slitasje indicator score."}
diff <- naturetypes_wide$var_7SE - 
        naturetypes_wide$var_7TK

@@

```{r}
path <- ifelse(dir == "C:", 
        "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")

ano <- sf::st_read(path,
                   layer = "ANO_SurveyPoint") %>%
  dplyr::filter(hovedoekosystem_punkt == "vaatmark")
```

@@

```{r, fig.cap="Distribution of slitasje scores in the ANO data"}
barplot(table(ano$slitasje), xlab="Sliatsje scores")
```

@@

Checking no that the new polygons have the area corresponding to the
proportion of the point that was part of the same main ecosystem:

@@

The new dataset only contains polygons.

@@

```{r, fig.cap="Slitasje scores."}
ggplot(temp, aes(x = Var1,
                 y = Freq))+
  geom_bar(stat="identity",
           fill="grey",
           colour = "black")+
  theme_bw(base_size = 12)+
  labs(x = "slitasje score",
       y = "Number of localities")
```

@@

```{r, fig.cap="Barplot show the contribution (number of localities) of different data sets to the slitasje indicator."}
temp <- as.data.frame(table(slitasje_data$origin))

@@

@@

1.  scale the indicators at the polygon level.

2.  take the simplified infrastructure index and, one class at the time,
    extract the indicator value that intersect with those cells, and
    extrapolate the mean of those values to all similar cell within a
    given region.

3.  calculate an indicator value for a region by doing a weighted
    average based o the relative area of each infrastructure class.

@@

```{r continuous-scale, fig.cap="Showing the conversion from ordinal to (pseudo)continous cale for the slitasje variable"}
ggplot(slitasje_data, aes(x = slitasje,
                          y = slitasje_percentage))+
  geom_point(size=6)+
  geom_line()+
  theme_bw(base_size = 16)
```

@@

```{r slitasje-normalise, fig.cap="Performing a non-linear break-point type normalisation of the slitasje variable."}
eaTools::ea_normalise(data = slitasje_data,
                      vector = "slitasje_percentage",
                      upper_reference_level = lower,
                      lower_reference_level = upper,
                      break_point = threshold,
                      plot=T,
                      reverse = T
                      )
```

@@

```{r}
HIA <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks/homogeneous_impact_areas.rds")
```

@@

```{r, echo=F}
# silently read cached file
HIA_regions <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_regions.rds")
```

@@

```{r, include=F}
#silently read cached file
all_stats <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats.rds")
```

@@

```{r, eval=F}
corrCheck <- st_intersection(slitasje_data, HIA_regions)
saveRDS(corrCheck, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")
```
```{r, include=F}
#Silently read cached data
corrCheck <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")
```

@@

natOpen_slitasje_extr <- ea_spread(indicator_data = natOpen,
                         indicator = i_2020,
                         regions = HIA_regions,
                         groups = region_HIF,
                         threshold = 100)
#saveRDS(wetlands_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
#saveRDS(seminat_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_slitasje_extr.rds")
#saveRDS(natOpen_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_slitasje_extr.rds")
```
```{r, include=F}
wetlands_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
seminat_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_slitasje_extr.rds")
natOpen_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_slitasje_extr.rds")
```

@@

```{r}
wetlands_slitasje_extr_trd <- sf::st_crop(wetlands_slitasje_extr, myBB)
seminat_slitasje_extr_trd <- sf::st_crop(seminat_slitasje_extr, myBB)
natOpen_slitasje_extr_trd <- sf::st_crop(natOpen_slitasje_extr, myBB)
```

@@

There is very little variation both between the ecosystems and the four classes in the modified infrastructure index. Note however, that for wetland and semi-natural areas we did not extrapolate values for the most urban areas, and that for naturally open areas we did not extrapolate to the least affected areas.

The next steps now are to

@@

Alternatively I can rasterize the indicator map, and mask it using a perfectly aligned ET map. This might be faster.

### Crop to ET
This is a very heavy operation.
```{r, eval=F}
mireET <- sf::st_read("R:/Kladd/Myr/Myrmodell.gdb")
```

```{r, eval=F}
wetlands_slitasje_extr_cropped <- sf::st_intersection(wetlands_slitasje_extr, mireET)
```

>>>>

```{r setup, include=FALSE}
library(knitr)
library(sf)
library(tmap)
library(ggplot2)
library(ggpubr)
library(DT)
library(units)
library(eaTools)
library(tidyverse)
library(tidyr)
knitr::opts_chunk$set(echo = TRUE)

# Set up conditional file paths
dir <- substr(getwd(), 1,2)

pData <- ifelse(dir == "C:", 
               "P:/41201785_okologisk_tilstand_2022_2023/data/",
               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")

@@

1.  Import [Nature type data](#NTM) data set (incl. GRUK) and [ANO data
    set](#ANO-import)

@@

6.  [Scale](#scale-slitasje-ind) the `slitasje` variable based on reference values.

@@

9. Confirm relationship between infrastructure index and indicator
    values to justify the extrapolation

10.  Sample 10 00 random points from an ecosystem delineation and extract indicator values and sd at those points.

11. Export indicator maps and regional extrapolated maps

## About the underlying data
The indicator uses a data set from a standardised field survey of nature types. You can read more about it [here](#naturtyper). I also import a data set called [ANO](#ANO-import)

### Representativity in time and space
The nature type mapping is not random and cannot be said to be area representative. 
The data is from 2018 to present. The data from one field season usually becomes available early the following year.

### Original units
The

@@

### UN standard

@@

```{r, eval=F}
naturetypes <- sf::st_read(dsn = paste0(pData, "Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"))
```

@@

```{r}
naturetypes <- readRDS(paste0(pData, "rds/naturtyper.rds"))
```

@@

```{r, fig.cap="7SE scores in the naturetype dataset", fig.width=3}
temp <- naturetypes_wide %>%
  group_by(var_7SE) %>%
  summarise(sum = n())

@@

```{r, fig.cap="7TK scores in the nature type dataset", fig.width=3}
temp <- naturetypes_wide %>%
  group_by(var_7TK) %>%
  summarise(sum = n())

@@

```{r slitasje-naturetype, fig.cap="Slitasje scores.", fig.width=3}
temp <- naturetypes_wide %>%
  group_by(slitasje) %>%
  summarise(sum = n())

@@

```{r, fig.cap="Counting the number of cases where 7TK or 7SE ws defining of the slitasje indicator score.", fig.width=3}
diff <- naturetypes_wide$var_7SE - 
        naturetypes_wide$var_7TK

@@

```{r}
ano <- sf::st_read(paste0(pData, "Naturovervaking_eksport.gdb"),
                   layer = "ANO_SurveyPoint") %>%
  dplyr::filter(hovedoekosystem_punkt == "vaatmark")
```

@@

```{r, fig.cap="Distribution of slitasje scores in the ANO data", fig.width=6}
barplot(table(ano$slitasje), xlab="Sliatsje scores", ylab = "Number of localities")
```

@@

Checking now that the new polygons have the area corresponding to the
proportion of the point that was part of the same main ecosystem:

@@

The new data set contains **only polygons**.

@@

#### Distribution of Slitasje scores
```{r, fig.cap="Slitasje scores (ANO Naturetype (and GRUK) data combined). The score for a locality equals the highest (worst) score of the related variables 7TK, 7SE, PRSL and PRTK."}
ggplot(temp, aes(x = Var1,
                 y = Freq))+
  geom_bar(stat="identity",
           fill="grey",
           colour = "black")+
  theme_bw(base_size = 12)+
  labs(x = "slitasje score",
       y = "Number of localities")
```

@@

```{r, fig.cap="Barplot show the contribution (number of localities) of different data sets to the slitasje indicator.", fig.width=4}
temp <- as.data.frame(table(slitasje_data$origin))

@@

The GIS layers are used to crop out marine areas, and to define accounting areas, respetively.

@@

1.  normalise the indicators at the polygon level.

2.  take the simplified infrastructure index and
    extract the indicator value that intersect with a given cell, and
    extrapolate the area weighted mean of those values to all similar cell within a
    given region (`ea_spread()`). Errors are calculated from bootstrapping.

3. Take the new area weighted indicator values for the homogeneous impact area (HIA) and match them with ecosystem occurrences of the same HIA. The errors are the same as for the HIAs.

3.  calculate an indicator value for a region (accounting area) by doing a weighted
    average based of the relative area of ecosystem occurrences. Errors shold be carried somehow, perhaps via weighted resampling

@@

```{r continuous-scale, fig.cap="Showing the conversion from ordinal to (pseudo)continous cale for the slitasje variable", fig.width=5}
ggplot(slitasje_data, aes(x = slitasje,
                          y = slitasje_percentage))+
  geom_point(size=6)+
  geom_line()+
  theme_bw(base_size = 16)
```

@@

```{r slitasje-normalise, fig.cap="Performing a non-linear break-point type normalisation of the slitasje variable.", fig.width=5}
eaTools::ea_normalise(data = slitasje_data,
                      vector = "slitasje_percentage",
                      upper_reference_level = lower,
                      lower_reference_level = upper,
                      break_point = threshold,
                      plot=T,
                      reverse = T
                      )
```

@@

```{r}
HIA <- readRDS(paste0(pData, "infrastrukturindeks/homogeneous_impact_areas.rds"))
```

@@

```{r, echo=F}
# silently read cached file

HIA_regions <- readRDS(paste0(pData, "cache/HIA_regions.rds"))
```

@@

```{r, include=F}
#silently read cached file
all_stats <- readRDS(paste0(pData, "cache/all_stats.rds"))
```

@@

```{r, eval=F}
corrCheck <- st_intersection(slitasje_data, HIA_regions)
saveRDS(corrCheck, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")
```
```{r, include=F}
#Silently read cached data
corrCheck <- readRDS(paste0(pData, "cache/corrCheck.rds"))
```

@@

natOpen_slitasje_extr <- ea_spread(indicator_data = natOpen,
                         indicator = i_2020,
                         regions = HIA_regions,
                         groups = region_HIF,
                         threshold = 100)
#saveRDS(wetlands_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
#saveRDS(seminat_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_slitasje_extr.rds")
#saveRDS(natOpen_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_slitasje_extr.rds")
```
```{r, include=F}

wetlands_slitasje_extr <- readRDS(paste0(pData, "cache/wetlands_slitasje_extr.rds"))
seminat_slitasje_extr  <- readRDS(paste0(pData, "cache/seminat_slitasje_extr.rds"))
natOpen_slitasje_extr  <- readRDS(paste0(pData, "cache/natOpen_slitasje_extr.rds"))
```

@@

```{r, warning=F}
wetlands_slitasje_extr_trd <- sf::st_crop(wetlands_slitasje_extr, myBB)
seminat_slitasje_extr_trd <- sf::st_crop(seminat_slitasje_extr, myBB)
natOpen_slitasje_extr_trd <- sf::st_crop(natOpen_slitasje_extr, myBB)
```

@@

There is some, but very little variation both between the ecosystems and the four classes in the modified infrastructure index. Note however, that for wetland and semi-natural areas we did not extrapolate values for the most urban areas, and that for naturally open areas we did not extrapolate to the least affected areas.

_These maps are easy to misunderstand, and they should therefor be communicated with caution._

Now, there are multiple way to go proceed. We have indicator maps (vector data), and we have HIA area, and we have the HIA with indicator data (also vector data).

Option 1:

@@

Step 1 above requires an st_interactions on two very large vactor data sets. This should be avoided.

Option 2:

1. rasterize the extrapolated indicator map, using the ET map as a tamplate, and

2. mask it using the perfectly aligned ET map.

This might be faster.

Option 3:

1. Draw n random point samples from the ET map, and

2. extract indicator values at those points (from the extrapolated indicator map (a stars object)).

This is perhaps even faster. I will try the last option, even though it will not provide spatailly explicit indicator values for the entire ecosystem delineation, like option 2 does. Perhaps I will revisit option 2 later.

### Ecosystem map
Importing the wetland map.

This is a quite heavy operation and should be done on a server.
```{r, eval=F}
mireET <- sf::st_read("/data/R/Kladd/Myr/Myrmodell.gdb")
```

This doesn't take that long.
```{r, eval=F}
mireET <- st_transform(mireET, 25833)
```

This take about 10 min. Not sure I need it.
```{r, eval=F}
mireET <- st_make_valid(mireET)
```

This takes a long!! time, and should be AVOIDED. 
```{r, eval=F}
wetlands_slitasje_extr_cropped <- sf::st_intersection(wetlands_slitasje_extr, mireET)
```

Instead I will draw some random point samples.
```{r}
wetland_points <- st_sample(mireET, 10000)
```

And intersect this with the extrapolated indicator values.
```{r}
wetland_points_match <- st_intersection(wetlands_slitasje_extr, wetland_points)
```

Create an empty grid, 50 x 50 m resolution
```{r}
grd = st_as_stars(st_bbox(wetlands_slitasje_extr), dx = 50, dy = 50, values = NA_real_)
```

Rasterize the indicator
```{r}
wetlands_slitasje_extr_rast <- st_rasterize(wetlands_slitasje_extr, grd)
```

And extract the values
```{r}
wetland_points_extracted <- st_extract(wetlands_slitasje_extr, wetland_points)
```
